\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-08-01 paper_OJA.tex
tweak the rendering. correct a typo on the SVI paragraph

%\documentclass[10pt,onecolumn,preprintnumbers,amsmath,amssymb,
%nofootinbib,superscriptaddress,aps,prd,numberedappendix]{openjournal}

\RequirePackage{rotating}
\PassOptionsToPackage{usenames,dvipsnames}{xcolour}
\documentclass[twocolumn,twocolappendix,nofootinbib]{openjournal}

% Add your own macros here:
\pdfoutput=1 %for arXiv submission
%\usepackage{amsmath,amssymb,amstext}
\usepackage{amsmath,amstext}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage{ae,aecompl}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[figure,figure*]{hypcap}
\usepackage{natbib}
\usepackage{url}
\usepackage{mdwlist}
%\usepackage{rotating}
\usepackage{multirow}
\urlstyle{same}

\usepackage{lineno}

%\documentclass[twocolumn,iop]{openjournal}
%\usepackage{newtxtext,newtxmath}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{graphicx}
%\usepackage{dcolumn}
%\usepackage{bm}
%\usepackage[dvipsnames]{xcolor}
%\usepackage{xspace}
%\usepackage{ulem}
%% The amssymb package provides various useful mathematical symbols

%\usepackage{fontawesome}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

%\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
%\usepackage[colorlinks = true,
%            linkcolor = xlinkcolor,
%            urlcolor  = xlinkcolor,
%            citecolor = xlinkcolor,
%            anchorcolor = xlinkcolor]{hyperref}

@@

%%
%% Python definition (c) 1998 Michael Weber
%% Additional definitions (2013) Alexis Dimitriadis
%% modified by me (should not have empty lines)
%%
\lstdefinelanguage{iPython}{
    morekeywords={access,and,break,class,continue,def,del,elif,else,except,exec,finally,for,from,global,if,import,in,is,lambda,not,or,pass,print,raise,return,try,while},%
    %
    % Built-ins
    morekeywords=[2]{abs,all,any,basestring,bin,bool,bytearray,callable,chr,classmethod,cmp,compile,complex,delattr,dict,dir,divmod,enumerate,eval,execfile,file,filter,float,format,frozenset,getattr,globals,hasattr,hash,help,hex,id,input,int,isinstance,issubclass,iter,len,list,locals,long,map,max,memoryview,min,next,object,oct,open,ord,pow,property,range,raw_input,reduce,reload,repr,reversed,round,set,setattr,slice,sorted,staticmethod,str,sum,super,tuple,type,unichr,unicode,vars,xrange,zip,apply,buffer,coerce,intern},%
    %
    sensitive=true,%
    morecomment=[l]\#,%
    morestring=[b]',%
    morestring=[b]",%
    %
    morestring=[s]{'''}{'''},% used for documentation text (mulitiline strings)
    morestring=[s]{"""}{"""},% added by Philipp Matthias Hahn
    %
    morestring=[s]{r'}{'},% `raw' strings
    morestring=[s]{r"}{"},%
    morestring=[s]{r'''}{'''},%
    morestring=[s]{r"""}{"""},%
    morestring=[s]{u'}{'},% unicode strings
    morestring=[s]{u"}{"},%
    morestring=[s]{u'''}{'''},%
    morestring=[s]{u"""}{"""},%
    %
    % {replace}{replacement}{lenght of replace}
    % *{-}{-}{1} will not replace in comments and so on
    literate=
    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
    {€}{{\EUR}}1 {£}{{\pounds}}1
    %
    {^}{{{\color{ipython_purple}\^{}}}}1
    {=}{{{\color{ipython_purple}=}}}1
    %
    {+}{{{\color{ipython_purple}+}}}1
    {*}{{{\color{ipython_purple}$^\ast$}}}1
    {/}{{{\color{ipython_purple}/}}}1
    %
    {+=}{{{+=}}}1
    {-=}{{{-=}}}1
    {*=}{{{$^\ast$=}}}1
    {/=}{{{/=}}}1,
    literate=
    *{-}{{{\color{ipython_purple}-}}}1
     {?}{{{\color{ipython_purple}?}}}1,
    %
    identifierstyle=\color{black}\ttfamily,
    commentstyle=\color{ipython_cyan}\ttfamily,
    stringstyle=\color{ipython_red}\ttfamily,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    %
    rulecolor=\color{ipython_frame},
    frame=none,
    frameround={t}{t}{t}{t},
    framexleftmargin=6mm,
    numbers=none,
    numberstyle=\tiny\color{halfgray},
    %
    %
    backgroundcolor=\color{ipython_bg},
    %   extendedchars=true,
    basicstyle=\scriptsize,
    keywordstyle=\color{ipython_green}\ttfamily,
}

@@

\newcommand{\FLan}[1]{{\color{cyan}FLan: #1}}

@@

\begin{document}
\journalinfo{The Open Journal of Astrophysics}
\submitted{submitted xxx, revised yyy, accepted zzz}

@@

\affiliation{$^1$Universit́e Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}

@@

We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
;in the early phases of Bayesian model development JAZ not sure what this means so I left it out

@@

In this context we have developed the open source \texttt{jax-cosmo} library \footnote{ \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper. The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.

@@

\FLa{Add description of problem and small demos}

@@

# Compute Fisher matrix of cosmological 
# parameters
F = - jax.hessian(likelihood)(cosmo)
\end{lstlisting}
%\end{minted}

@@

-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}

@@

to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to

@@

Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to use these schema to ease the model development in its early phase.
%
\section{Discussion}
\label{sec-discussion}
%
The key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie. about $1\%$). So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.

@@

\FLa{mention something about scaling with parallel MCMC}

@@

with seed(rng_seed=42):
    data, P, C = fiducial_model()
%\end{minted}
\end{lstlisting}

@@

mcmc.run(jax.random.PRNGKey(42))
\end{lstlisting}
%\end{minted}
More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.
%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%\section*{Bibliography}
\bibliographystyle{elsarticle-harv} 
\typeout{}
\bibliography{refs}

>>>>

\PassOptionsToPackage{usenames,dvipsnames}{xcolour}
\documentclass[twocolumn,twocolappendix,nofootinbib,iop]{openjournal}

% Add your own macros here:
\pdfoutput=1 %for arXiv submission
\usepackage{amsmath,amssymb,amstext}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage{ae,aecompl}
\usepackage[utf8]{inputenc}
%\usepackage{hyperref}
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage[figure,figure*]{hypcap}
\usepackage{natbib}
\usepackage{url}
\usepackage{mdwlist}
\usepackage{multirow}
\urlstyle{same}
\usepackage{lineno}
\usepackage{fontawesome}

@@

%%
%% Python definition (c) 1998 Michael Weber
%% Additional definitions (2013) Alexis Dimitriadis
%% modified by me (should not have empty lines)
%%
\lstdefinelanguage{iPython}{
    morekeywords={access,and,break,class,continue,def,del,elif,else,except,exec,finally,for,from,global,if,import,in,is,lambda,not,or,pass,print,raise,return,try,while},%
    %
    % Built-ins
    morekeywords=[2]{abs,all,any,basestring,bin,bool,bytearray,callable,chr,classmethod,cmp,compile,complex,delattr,dict,dir,divmod,enumerate,eval,execfile,file,filter,float,format,frozenset,getattr,globals,hasattr,hash,help,hex,id,input,int,isinstance,issubclass,iter,len,list,locals,long,map,max,memoryview,min,next,object,oct,open,ord,pow,property,range,raw_input,reduce,reload,repr,reversed,round,set,setattr,slice,sorted,staticmethod,str,sum,super,tuple,type,unichr,unicode,vars,xrange,zip,apply,buffer,coerce,intern},%
    %
    sensitive=true,%
    morecomment=[l]\#,%
    morestring=[b]',%
    morestring=[b]",%
    %
    morestring=[s]{'''}{'''},% used for documentation text (mulitiline strings)
    morestring=[s]{"""}{"""},% added by Philipp Matthias Hahn
    %
    morestring=[s]{r'}{'},% `raw' strings
    morestring=[s]{r"}{"},%
    morestring=[s]{r'''}{'''},%
    morestring=[s]{r"""}{"""},%
    morestring=[s]{u'}{'},% unicode strings
    morestring=[s]{u"}{"},%
    morestring=[s]{u'''}{'''},%
    morestring=[s]{u"""}{"""},%
    %
    % {replace}{replacement}{lenght of replace}
    % *{-}{-}{1} will not replace in comments and so on
    literate=
    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
    {€}{{\EUR}}1 {£}{{\pounds}}1
    %
    {^}{{{\color{ipython_purple}\^{}}}}1
    {=}{{{\color{ipython_purple}=}}}1
    %
    {+}{{{\color{ipython_purple}+}}}1
    {-}{{{\color{ipython_purple}-}}}1
    {*}{{{\color{ipython_purple}$^\ast$}}}1
    {/}{{{\color{ipython_purple}/}}}1
    %
    {+=}{{{+=}}}1
    {-=}{{{-=}}}1
    {*=}{{{$^\ast$=}}}1
    {/=}{{{/=}}}1,
    literate=
    *{-}{{{\color{ipython_purple}-}}}1
     {?}{{{\color{ipython_purple}?}}}1,
    %
    identifierstyle=\color{black}\ttfamily,
    commentstyle=\color{ipython_cyan}\ttfamily,
    stringstyle=\color{ipython_red}\ttfamily,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    %
    rulecolor=\color{ipython_frame},
    frameround={t}{t}{t}{t},
    numbers=none,
    numberstyle=\tiny\color{halfgray},
    %
    %
    backgroundcolor=\color{ipython_bg},
    %   extendedchars=true,
    %basicstyle=\scriptsize,
    basicstyle=\ttfamily\footnotesize,
    columns=fullflexible,
    keywordstyle=\color{ipython_green}\ttfamily,
}

@@

%%% color Comments 
\newcommand{\FrL}[1]{{\color{cyan}FL: #1}}

@@

\begin{document}
\journalinfo{The Open Journal of Astrophysics}
%\submitted{submitted xxx, revised yyy, accepted zzz}

@@

\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}

@@

We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
{\color{red}
;in the early phases of Bayesian model development JAZ not sure what this means so I left it out}

@@

In this context we have developed the open source \texttt{jax-cosmo} library \footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper. The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.

@@

\FrL{Add description of problem and small demos}

@@

# Compute Fisher matrix of cosmological 
# parameters
F = -jax.hessian(likelihood)(cosmo)
\end{lstlisting}
%\end{minted}

@@

-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}

@@

to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}

@@

Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
%
\section{Discussion}
\label{sec-discussion}
%
The key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie. about $1\%$). So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.

@@

\FrL{mention something about scaling with parallel MCMC}

@@

with seed(rng_seed=42):
    data, P, C = fiducial_model()
\end{lstlisting}
%\end{minted}

@@

mcmc.run(jax.random.PRNGKey(42))
\end{lstlisting}
%\end{minted}
More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.
%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%\section*{Bibliography}
%\bibliographystyle{elsarticle-harv} 
\bibliographystyle{aa} 
\typeout{}
\bibliography{refs}
